---
title: "About me"
# subtitle: "Why you'd want to go on a date with me"
comments: false
author: "Hansjörg Plieninger"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

## About

My name is Hansjörg Plieninger, I am a postdoctoral researcher in the psychology department at the University of Mannheim, Germany. I love data science and I love R. From an academic perspective, I am an expert in psychometrics, item response theory, structural equation modeling, and multilevel modeling; more specifically, I use this machinery to investigate response styles (e.g., tendency to give extreme responses). From a data-science perspective, I am an expert in dealing with survey data, Likert-type data and the like.

This blog is about ...

* things I learned to make it simpler for other people to learn the same thing quicker,
* things I learned to make it simpler for me to do the same thing quicker next time,
* things I love to make it simpler for others to love them, too,
* things I do to make it simpler for you to get in touch with me.

And the motto of this blog comes from one of my favourite R packages, **fortunes**:

```{r}
fortunes::fortune("done it.")
```
