---
title: Hello World
author: Hansjörg Plieninger
date: '2018-03-01'
slug: hello-world
categories: []
tags: []
description: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

Hello. My name is Hansjörg, I am a postdoctoral researcher in the psychology department at the University of Mannheim, Germany. I want to start blogging today, because I always wanted to, and because one of my R heroes Yihui Xie made this possible through his R package **blogdown**. Thank you.

```{r echo=FALSE}
blogdown::shortcode('tweet', '943194437076897792')
```

This blog, as of today, serves the following purposes:

* Blog about things I learned to make it easier for other people to learn the same thing quicker.
* Blog about things I learned to make it easier for me to do the same thing quicker next time.
* Blog about things I love to make it easier for others to love them, too.
* Blog about things I do to make it easier for you to get in touch with me.

```{r}
fortunes::fortune("done it.")
```
