---
title: 'Hello World'
author: 'Hansjörg Plieninger'
date: '2018-02-01'
slug: hello-world
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
```

Hello, my name is Hansjörg, and I am a postdoctoral researcher in the psychology department at the University of Mannheim, Germany. I want to start blogging today, because I always wanted to, and because one of my R heroes Yihui Xie made this possible through his R package **blogdown**:

```{r echo=FALSE}
blogdown::shortcode('tweet', '943194437076897792')
```

This blog is about ...

* things I learned, to make it simpler for other people to learn the same thing quicker,
* things I learned,  to make it simpler for me to do the same thing quicker next time,
* things I love, to make it simpler for others to love them, too,
* things I do, to make it simpler for you to get in touch with me.

And the motto of this blog comes from one of my favorite R packages, **fortunes**:

```{r}
fortunes::fortune("done it.")
```
