---
title: 'Font Embedding for LaTeX and R users'
subtitle: 'Use cairo_pdf()'
author: 'Hansjörg Plieninger'
date: '2018-02-15'
slug: font-embedding-for-latex-and-r-users
categories:
  - R
tags:
  - R
  - LaTeX
  - pdf
  - note-to-myself
output:
  blogdown::html_page:
    toc: true
---

Recently, I sent my dissertation as a PDF file to a copy shop and got an email back that I had not embedded all fonts and that they won't print it for me. What? So instead of celebrating the submission, I had to search online for *pdf latex "font embedding"*, and this blog post is a summary of that afternoon.

## What is Font Embedding

In order to display or print a PDF, the system has to know which fonts it should use.

* If the font used in the PDF is available on the system, it is able to handle it correctly (e.g., Arial).
* If the font used in the PDF is not available on the system but is embedded in the PDF, the system is able to handle it correctly.
* If the font is not available and not embedded, the system will guess what font to use for display/print and this may or may not work. This [link](http://wiki.opf-labs.org/download/attachments/101613571/fonts_notembedded.jpg?version=1&modificationDate=1430223649000) shows an extreme example of what might happen.

Embedding fonts is probably not crucial for non-exotic fonts on non-exotic devices/printers.
However, if you want to make sure that your PDF is correctly displayed or printed on any possible device, fonts should be embedded.
Such devices could be, for example, a printer of a copy shop, an exotic device of a student, or a person with an otherwise different setup (e.g., someone in the year 2076 in Tanzania).

## Problem: R does not embed fonts

```{r, eval = F}
pdf()
plot(cars)
dev.off()
```

R does not embed fonts when producing a PDF via `pdf()` as above. Thus, if you compile a PDF with LaTeX and include R plots (i.e., PDFs), LaTeX cannot embed the fonts of the plots, and then you finally have a PDF with some fonts not embedded. LaTeX itself seems to embed all fonts automatically, so if you don't include any plots, you don't have to worry.

You can check whether fonts are embedded with most PDF readers and something along the lines of Properties -> Fonts.

## Solution in R: Use cairo_pdf()

If you wish to embed the fonts of your R plots, you have the following options.

* Do not produce vector graphics via `pdf()` but images via `png()`, for example. This is probably not the desired solution because of quality, but it circumvents the problem of font embedding.
* Produce vector graphics not via `pdf()` but via `cairo_pdf()`. This will embed the fonts automatically, and LaTeX will subsequently embed these fonts, too. This is the workflow that I recommend.
* Postscript may also be an alternative (e.g., `cairo_ps()`), but I don't know much about postscript.
* You may also embed fonts post hoc. R has a function `embedFonts()`, and the package **extrafont** provides a wrapper, for example, `extrafont::embed_fonts("f1.pdf", outfile="f1_embedded.pdf")`.  
BTW: You may use any font (available on your system) in any R plot with the **extrafont** package.


## Solution With Completed PDF

If you already have a PDF with unembedded fonts and want to embed the fonts post hoc, things are different and may be a bit more difficult. What worked for me in a one or two test cases are the following solutions.

* Use ghostscript via `embedFonts()` or the **extrafont** package as stated above.
* You may also call ghostscript directly (rather than from R) as described in a [blog post by Karl Rupp](https://www.karlrupp.net/2016/01/embed-all-fonts-in-pdfs-latex-pdflatex/), close to the end of the post. What worked for me is the following command in a Windows Command Prompt:
	
```{r, eval = F}
gswin64c -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -dPDFSETTINGS=/prepress -dEmbedAllFonts=true -sOutputFile=C:\...\my_plot_embedded.pdf -f C:\...\my_plot.pdf
```

These solutions are certainly dependent on the fact that you have the fonts, which you want to embed, installed on your computer.
