---
title: 'Working With Different Versions of an R Package'
author: 'Hansjörg Plieninger'
date: '2018-03-06'
summary: 'Recently, I had to install an older version of an R package, because a function was deprecated that I wanted to use. I wanted to install the old version in addition to and not instead of the new version and I came up with the following solution.'
categories: []
tags:
  - r
  - note-to-myself
slug: multiple-package-versions
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", out.width = "80%")
```

Recently, I had to install an older version of an R package, because a function was deprecated that I wanted to use.
I wanted to install the old version in addition to and not instead of the new version.

*This post has been updated in April 2020.
Previously, `install_version()` and `install_github()` had no `lib` argument which made it necessary to use a workaround via the withr package to install to a non-standard library.
Today, these functions pass `lib` to `install.packages()`, which makes this whole endeavor a lot easier.*

## Installing Packages to a Custom Location

```{r}
library("remotes")
library("fs")
```

In addition to my standard Windows library, I created two new folders to store packages, one for older packages and one for devel versions.

```{r}
dev_lib <- path_home_r("R/win-library/dev-versions/")
old_lib <- path_home_r("R/win-library/old-versions/")

# dir_create(dev_lib)
# dir_create(old_lib)
```

The aim was to have a folder structure similar to the one shown below:

```
C:/Users/hp/Documents/R
└── win-library
    ├── 3.5
    ├── 3.6
    |   ├── ...
    |   ├── dplyr
    |   ├── ...
    │   ├── MplusAutomation
    |   └── ...
    ├── dev-versions
    │   └── dplyr
    └── old-versions
        └── MplusAutomation
```

## Installing Development Packages

The [remotes](https://remotes.r-lib.org/) package has many functions to install packages from various sources.
The functions in remotes used to live in the devtools package and can still be used via devtools.

Probably the function that is used most is `install_github()`, which I use here to install the current version of the dplyr package from https://github.com/tidyverse/dplyr/.

```{r, eval = FALSE}
install_github("tidyverse/dplyr", lib = dev_lib, upgrade = "always")
```

Now, I can either load the CRAN version or the development version.
As of this writing, the new function `relocate()` is not yet part of the CRAN version such that the first attempt fails.

```{r, message = FALSE, warning = FALSE}
head(iris)

library("dplyr")
packageVersion("dplyr")
try(relocate(iris, Species))

detach("package:dplyr")

library("dplyr", lib.loc = dev_lib)
packageVersion("dplyr")
head( relocate(iris, Species) )
```

## Installing Outdated Packages

`remotes::install_version()` allows to install an outdated version of a package automatically without having to manually download the archive.

```{r, eval = FALSE}
install_version("MplusAutomation", version = "0.7", lib = old_lib)
```

Now, I can either load the CRAN version or the development version.
In the current CRAN version, the function `extractModelSummaries()` is no longer available such that the first attempt fails.

```{r, eval = TRUE, warning=FALSE}
suppressPackageStartupMessages(
    library("MplusAutomation")
)
packageVersion("MplusAutomation")

MplusAutomation::extractModelSummaries()

detach("package:MplusAutomation")

library("MplusAutomation", lib.loc = old_lib)
packageVersion("MplusAutomation")

# works with a valid target:
# MplusAutomation::extractModelSummaries(target = NULL)
```

## Special cases

As noted above, the package [withr](http://withr.r-lib.org/) can give you further control if that is necessary.
In the original version of this blog post, I used `with_libpaths()`, since `install_github()` did not accept a `lib` argument in the past.

```{r, eval = FALSE}
library("withr")
with_libpaths(new = dev_lib,
              code = install_github("tidyverse/dplyr",
                                    dependencies = FALSE))
```

## Reproducibility via a Project-Specific Library

Sometimes, you want more control.
You may want to specify exactly the version of an R package that should be used (e.g., dplyr 0.8.0.1).
This is the time for a dependency management system that allows you to have project-specific libraries.
In R, this is best managed by means of the [packrat](https://rstudio.github.io/packrat/) package or its successor, the [renv](https://rstudio.github.io/renv/) package.
