---
title: "Working with different versions of an R package"
author: "Hansjörg"
date: 2018-03-06
categories: ["R"]
tags: ["R"]
slug: working-with-different-package-versions
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")


library(cli)

options(crayon.enabled = TRUE)
options(crayon.colors = 256)
crayon::num_colors(forget = TRUE)
```

## Motivation

Recently, I had to install an older version of an R package, because a function was deprecated that I wanted to use.
I wanted to install the old version in addition to and not instead
of the new version and I came up with the following solution.

## Installing Packages to a Different Location

```{r, eval = F}
library(devtools)
library(httr)
library(curl)
```

In addtion to my custom Windows library, I created two new folders to store packages in, one for older packages and one for devel versions.

```{r, eval = T}
dev_lib <- file.path("~/../Documents/R/win-library/dev-versions/")
old_lib <- file.path("~/../Documents/R/win-library/old-versions/")

list.dirs("~/../Documents/R/win-library", recursive = FALSE)
```



```{r tree, render=ansistrings:::knit_print.html, eval = F, include = F}

# ```{r tree, render=ansistrings:::knit_print.html, echo = F}
# ```{r tree, render=function(x, ...) ansistrings:::knit_print.html(x, zoom = 10), echo = F}

dat <- data.frame(
  stringsAsFactors = FALSE,
  package = c(
      "R",
      "win.library",
      "3.4",
      "dev.versions",
      "old.versions",
      "ggplot2",
      "...",
      "MplusAutomation"
      ),
  dependencies = I(list(
      "win.library",
      c("3.4", "dev.versions", "old.versions"),
      c("...", "ggplot2", "...", "MplusAutomation", "..."),
      character(0),
      character(0),
      character(0),
      character(0),
      character(0)
      ))
  )
tree(dat, root = "R")
```

Then, I can install packages in these folders using custom arguments:
 
```{r, eval = FALSE}
install.packages("~/../Downloads/MplusAutomation_0.7.tar.gz",
                 lib = old_lib, repos = NULL)
```

## Installing Packages to a Different Location Using devtools

`devtools::install_version()` allows to install a specific version of a package automatically without having to manually download the achive.
However, you cannot use a custom path, but `withr::with_libpaths()` can take care of that. Here, for example, I install an old version of **MplusAutomation** and the devel version of **ggplot2**.

```{r, eval = F}
withr::with_libpaths(new = old_lib,
                     code = install_version("MplusAutomation",
                                            version = "0.7",
                                            dependencies = FALSE))
``` 

```{r, eval = F}
withr::with_libpaths(new = dev_lib,
                     code = install_github("tidyverse/ggplot2",
                                           dependencies = FALSE))
```

Note, that I loaded **devtools**, **httr**, and **curl** above, because they need to be available inside `with_libpaths()`. Furthermore, I used `dependencies = FALSE`, because I already have the package dependencies in my default folder `"/win.library/3.4/"` and do not want them to be installed in the custom folders as well.

## Loading Packages

When I load **MplusAutomation** (version 0.7-1), the function `extractModelSummaries()` is not available, and the call returns an error.

```{r, eval = T, warning=FALSE}
suppressPackageStartupMessages(
    library(MplusAutomation)
    )

# doesn't work:
MplusAutomation::extractModelSummaries()
```

However, when I load the old version 0.7, I can use this function as I did in previous years.

```{r, eval = FALSE}
library(MplusAutomation, lib.loc = old_lib)

# works:
MplusAutomation::extractModelSummaries()
```
